---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Installation and Setup

```{r}
#BiocManager::install("muscData")
#BiocManager::install("muscat")

```

```{r}
library(muscData)
library(Seurat)
library(muscat)
library(tidyverse)

sce <- Kang18_8vs8()
```

---

## Single Cell Experiment

Single cell experiment objects are the R equivalent of an anndata object and store all of the data for a single cell experiment in a single data structure.

This will return the basic dimensions of the dataset in the form of genes x cells:

```{r}
dim(sce)
```

---

## Three main components of the object

1. **Assays** (count matrices)
2. **colData** (cell-level metadata)
3. **rowData** (gene-level metadata)

```{r}

counts(sce)
```

```{r}

head(colData(sce))
```

```{r}

head(rowData(sce))
```

---

## Exploring Cell Metadata

```{r}
colData(sce)
```

The `ind` column contains the ID of each individual sample. How many unique samples are there?

```{r}
length(unique(colData(sce)$ind))

```

The `stim` column indicates whether cells were stimulated with interferon-β or left as controls. How many cells are in each condition?

```{r}
table(sce$stim)

```

Create a cross-tabulation showing the number of cells for each individual in each condition:

```{r}
table(Individual = sce$ind, Condition = sce$stim)

```

The `cell` column contains cell type annotations. Which cell type has the most cells?

```{r}
cell_counts <- table(sce$cell)
sort(cell_counts, decreasing = TRUE)

```

---

## Exploring Gene Metadata

```{r}
head(rowData(sce))
```

The gene metadata contains Ensembl IDs and gene symbols. Find the gene symbol for "ENSG00000087086":

```{r}
rowData(sce)[rowData(sce)$ENSEMBL == "ENSG00000087086", ]

```

How many genes have the symbol "ISG15"?

```{r}
sum(rowData(sce)$SYMBOL == "ISG15")

```

---

## Subsetting SingleCellExperiment Objects

SCE objects use `sce[genes, cells]` syntax for subsetting.

```{r}
#!is.na(sce$cell) & sce$cell == "B cells"
sce_bcells <- sce[, !is.na(sce$cell) & sce$cell == "B cells"]
dim(sce_bcells)
```

```{r}

sce_bcells_stim <- sce_bcells[, !is.na(sce_bcells$stim) & sce_bcells$stim == "stim"]
dim(sce_bcells_stim)
```

Create a subset containing only CD14+ Monocytes from individual "101":

```{r}


mono_filter <- !is.na(sce$cell) & sce$cell == "CD14+ Monocytes" & 
               !is.na(sce$ind) & sce$ind == "101"

sce_mono_101 <- sce[, mono_filter]
dim(sce_mono_101)

```

Create a subset containing cells from individuals "101" and "107", and only genes that start with "ISG":

```{r}

cell_filter <- !is.na(sce$ind) & sce$ind %in% c("101", "107")
gene_filter <- !is.na(rowData(sce)$SYMBOL) & grepl("^ISG", rowData(sce)$SYMBOL)


sce_subset <- sce[gene_filter, cell_filter]
dim(sce_subset)


```

---

## Quality Control Metrics

```{r}

lib_sizes <- colSums(counts(sce))
summary(lib_sizes)
```

```{r}

n_genes <- colSums(counts(sce) > 0)
summary(n_genes)
```

```{r}

sce$lib_size <- lib_sizes
sce$n_genes <- n_genes
colData(sce)
```


## Preparing for Pseudobulk Analysis

For pseudobulk analysis, we aggregate cells by individual and condition.

```{r}

sce$sample_id <- paste(sce$ind, sce$stim, sep = "_")


```

```{r}

table(sce$sample_id)
```

For CD14+ Monocytes, how many cells contribute to each pseudobulk sample?

```{r}
sce_mono <- sce[, !is.na(sce$cell) & sce$cell == "CD14+ Monocytes"]
cells_per_sample <- table(sce_mono$sample_id)
cells_per_sample
```


Use the aggregateData function from muscat to generate matrices where you've added
up the counts from each cell population to a single count per sample. Pseudobulking
has been shown to mitigate many of the biases when trying to compare expression 
profiles of individual cells of the same type against individual cells. 

You will want to use the `cell` and `sample_id` column to group and count cells
```{r}
BiocManager::install("muscat")
library(muscat)
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cell", "sample_id"))
```
  
 
Choose just one of the cell populations at random, and extract out the counts
using assay(sce_object, "name_in_column")
```{r}

b.cells.cts <- assay(pb, "B cells")

```
 
 
```{r}
coldata <- data.frame(condition = colnames(b.cells.cts) %>% str_split_i("_", 2))
rownames(coldata) <- colnames(b.cells.cts)
coldata
```
 
```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = b.cells.cts,
                              colData = coldata,
                              design = ~ condition)


```


```{r}
dds <- DESeq(dds)
res <- results(dds)
resOrdered <- res[order(res$pvalue),]
```

```{r}
resOrdered %>% as_tibble(rownames='id')
```

```{r}
plotCounts(dds, gene="ISG15", intgroup="condition")
```
 
```{r}
seurat <- CreateSeuratObject(counts = counts(sce), meta.data = as.data.frame(colData(sce)))

```

```{r}
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, pattern = "^Mt")
VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat, selection.method = "vst", nfeatures = 2000)

all.genes <- rownames(seurat)
seurat <- ScaleData(seurat, features = all.genes)

seurat <- RunPCA(seurat, features = VariableFeatures(object = seurat))
ElbowPlot(seurat)
```

```{r}
seurat <- FindNeighbors(seurat, dims = 1:20)
seurat <- FindClusters(seurat, resolution = 0.5)
seurat <- RunUMAP(seurat, dims = 1:20)
DimPlot(seurat, reduction = "umap")
```

```{r, fig.width=16, fig.height=8}
DimPlot(seurat, reduction = "umap", group.by=c('stim', 'seurat_clusters'), label=T)
```


```{r}
seurat.markers <- FindAllMarkers(seurat)
```

```{r, fig.height=10, fig.width=10}
seurat.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10
DoHeatmap(seurat, features = top10$gene) + NoLegend()
```