---
title: "Human Liver Cirrhosis scRNA-seq — Part 1: Loading, QC & Integration"
subtitle: "Ramachandran et al. 2019 | GSE136103"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

## Learning Objectives

By the end of this exercise you will be able to:

1. Load 10X Genomics count matrices into Seurat and understand the object structure
2. Calculate and interpret standard single-cell QC metrics
3. Apply adaptive MAD-based filtering rather than arbitrary hard cutoffs
4. Detect and remove putative doublets on a per-sample basis
5. Merge multiple samples and understand why donor batch correction is necessary
6. Run Harmony integration and evaluate whether it has worked

**Estimated time:** 2–3 hours

---

## Background

Ramachandran et al. (2019, *Nature*) profiled the transcriptomes of more than 100,000 single
cells from healthy and cirrhotic human liver. Their key finding was a novel
**TREM2+CD9+ macrophage subpopulation** that expands in liver fibrosis. They also identified
**ACKR1+** and **PLVAP+** endothelial cells restricted to the fibrotic scar.

The dataset (GEO accession **GSE136103**) contains cells from:

- **5 healthy donors** and **5 cirrhotic donors** (causes: NAFLD, alcohol, PBC)
- Two sorted fractions per donor: **CD45+** (immune cells) and **CD45−** (endothelial,
  mesenchymal cells)
- Some donors have their CD45− fraction split across **two sequencing lanes (A and B)**

In Part 1 we focus entirely on producing a clean, well-annotated Seurat object ready for
clustering and biological interpretation in Part 2.

---

## 0. Libraries

▶ **Provided** — run this chunk as-is. Most of these packages should be pre-installed
on the SCC.

```{r libraries}
library(Seurat)       # core single-cell analysis framework
library(tidyverse)    # data wrangling (dplyr, tidyr) and plotting (ggplot2)
library(patchwork)    # arrange multiple ggplot panels side by side
library(scuttle)      # MAD-based QC filtering (Bioconductor)
library(scDblFinder)  # doublet detection (Bioconductor)
library(harmony)      # batch integration

set.seed(42)  # fix random seed for reproducibility
```

```{r}
BiocManager::install("scDblFinder")
```

---

## 1. Understanding the Data Structure

Before writing any code, let us understand what files we have and how they are organised.

Each sample has three files that together define the count matrix:

| File | Contents |
|------|----------|
| `barcodes.tsv.gz` | One cell barcode per row — each barcode is one cell |
| `features.tsv.gz` | One gene per row |
| `matrix.mtx.gz` | The counts in sparse format: (gene index, cell index, count) |

After reorganising the downloaded files, each sample lives in its own subdirectory:

```
data/
├── GSM4041150_healthy1_cd45+/
│   ├── barcodes.tsv.gz
│   ├── features.tsv.gz
│   └── matrix.mtx.gz
├── GSM4041151_healthy1_cd45-A/
│   └── ...
```

You may find the data on the cluster at /projectnb/bf530/materials/lab11-data/data/

> **Question 1:** Each donor has a `cd45+` and a `cd45-` fraction. What cell types would
> you expect to find in each? Why might the authors have sorted cells rather than loading
> an unsorted suspension?
>
> *Your answer:*

▶ **Provided** — set your data path and list the sample directories.

```{r data-dir}
# Set this to the folder listed above
data_dir <- "/projectnb/bf530/materials/lab11-data/data/"   # <-- adjust this path if needed

# list.dirs() returns the path of every subdirectory inside data_dir
# recursive = FALSE means we only go one level deep (we don't want subdirectories
# of subdirectories)
# full.names = TRUE means we get the complete path, not just the folder name
sample_dirs  <- list.dirs(data_dir, recursive = FALSE, full.names = TRUE)

# basename() strips the directory path and keeps only the final folder name
# e.g. "data/GSM4041150_healthy1_cd45+" becomes "GSM4041150_healthy1_cd45+"
sample_names <- basename(sample_dirs)

cat("Number of samples found:", length(sample_names), "\n\n")
print(sample_names)
```

### 1.1 Sample Metadata

▶ **Provided** — There's metadata table so you can join sample information without
parsing filenames.

```{r metadata}
sample_meta <- read.csv("samplesheet.tsv", sep = "\t", stringsAsFactors = FALSE)
print(sample_meta)

# or equivalently: sample_meta <- read.delim("samplesheet.tsv", stringsAsFactors = FALSE)

```

| Column | Meaning |
|--------|---------|
| `sample_name` | Directory name |
| `donor` | Donor ID (e.g. healthy1, cirrhotic3) |
| `condition` | healthy or cirrhotic |
| `fraction` | cd45pos or cd45neg |
| `etiology` | healthy, NAFLD, alcohol, or PBC |
| `lane` | single, A, or B |

---

## 2. Loading a Single Sample

We always load one sample by hand first to understand what we are creating before
automating it across all samples. It is much easier to debug a problem with one object
than to realise something went wrong after processing twenty of them.

▶ **Provided** — load the first sample and inspect the raw count matrix.

```{r load-one}
path_one   <- sample_dirs[1]
sample_one <- sample_names[1]

cat("Loading:", sample_one, "\n\n")

# Read10X() is Seurat's function for reading 10X Genomics output.
# It reads all three files in the directory simultaneously and assembles them
# into a single sparse matrix where rows are genes and columns are cell barcodes.
# It handles the .gz compression automatically.
counts_one <- Read10X(data.dir = path_one)

cat("Dimensions (genes x cells):", dim(counts_one), "\n")
```

```{r inspect-counts}
# The . characters represent zeros -- this is how R displays sparse matrices.
# Most entries are zero because each cell only expresses a fraction of all genes.
# This sparsity is normal and expected in single-cell data.
counts_one[1:6, 1:6]
```

▶ **Provided** — create a Seurat object and explore its structure.

```{r create-seurat-one}
# CreateSeuratObject() wraps the raw count matrix in a Seurat object,
# which is a structured container that will hold our data, metadata, and
# analysis results throughout the exercise.
#
# min.cells = 3 : drop any gene that appears in fewer than 3 cells.
#   These very sparse genes add noise and memory usage without contributing
#   useful information for identifying cell types.
#
# min.features = 200 : drop any cell (barcode) with fewer than 200 genes detected.
#   Barcodes with very few genes are almost certainly empty droplets that captured
#   only a small amount of ambient RNA, not a real cell.
so_one <- CreateSeuratObject(
  counts       = counts_one,
  project      = sample_one,
  min.cells    = 3,
  min.features = 200
)

so_one  # printing a Seurat object shows a brief summary
```

```{r explore-seurat}
# nrow() and ncol() work on Seurat objects: rows = genes, columns = cells
cat("Genes:", nrow(so_one), "\n")
cat("Cells:", ncol(so_one), "\n\n")

# colnames() returns the cell barcodes (one per cell)
cat("First 6 cell barcodes:\n")
head(colnames(so_one))

# rownames() returns the gene names
cat("\nFirst 6 gene names:\n")
head(rownames(so_one))

# @meta.data is a data frame with one row per cell.
# It stores per-cell information: QC metrics, cluster labels, etc.
# nCount_RNA and nFeature_RNA are added automatically by CreateSeuratObject.
cat("\nMetadata (first 6 rows):\n")
head(so_one@meta.data)
```

The `@meta.data` slot is a data frame with one row per cell. `CreateSeuratObject`
automatically calculates `nCount_RNA` (total UMIs) and `nFeature_RNA` (genes detected).
We will add more columns throughout this exercise.

---

## 3. Loading All Samples

Now we scale up from one sample to all samples. Rather than copying and pasting the
loading code twenty times (error-prone and hard to maintain), we wrap it in a function
and apply it to every sample automatically.

**The pattern used here — `lapply()` and `mapply()`** — is fundamental in R and worth
understanding:

- `lapply(X, FUN)` applies a function `FUN` to each element of a list `X` and returns
  a list of results. 
- `mapply(FUN, arg1, arg2)` is similar but lets you vary multiple arguments at once —
  here we need to pass both a file path and a sample name for each sample.

Using these instead of a `for` loop means each sample is processed independently,
the results are automatically collected into a list, and the code is shorter and
less likely to contain indexing mistakes.

▶ **Provided** — this function loads one sample and attaches metadata. Read it carefully
before running.

```{r load-all}
# This function encapsulates everything needed to load one sample.
# Writing it as a function means:
# 1. We can test it on a single sample before running it on all of them
# 2. The code only needs to be written once -- no copy-paste errors
# 3. It is easy to add a step (e.g. extra QC metric) in one place
load_one_sample <- function(path, sname, metadata_table) {
  
  counts <- Read10X(data.dir = path)
  
  so <- CreateSeuratObject(
    counts       = counts,
    project      = sname,
    min.cells    = 3,
    min.features = 200
  )
  
  # filter() from dplyr finds the row in sample_meta that matches this sample.
  # We then pull each column value out and attach it directly to the Seurat object.
  # Attaching metadata now (before merging) means we never lose track of which
  # cells came from which donor, condition, or fraction.
  meta_row <- filter(metadata_table, sample_name == sname)
  
  so$sample    <- sname
  so$donor     <- meta_row$donor
  so$condition <- meta_row$condition
  so$fraction  <- meta_row$fraction
  so$etiology  <- meta_row$etiology
  so$lane      <- meta_row$lane
  
  return(so)
}

# mapply() applies load_one_sample() to every sample simultaneously.
# It pairs up sample_dirs[1] with sample_names[1], sample_dirs[2] with
# sample_names[2], and so on. metadata_table is the same for every call
# so it goes in MoreArgs.
# SIMPLIFY = FALSE ensures the result stays as a list of Seurat objects
# rather than being collapsed into a simpler structure.
seurat_list <- mapply(
  FUN         = load_one_sample,
  path        = sample_dirs,
  sname       = sample_names,
  MoreArgs    = list(metadata_table = sample_meta),
  SIMPLIFY    = FALSE
)

# Give each list element the name of its sample for easy access later
names(seurat_list) <- sample_names

# sapply() is like lapply() but tries to simplify the result.
# Here ncol() returns a single number per sample, so sapply() gives
# us a named numeric vector -- convenient for a quick overview.
cat("Cells loaded per sample:\n")
sapply(seurat_list, ncol)
```

Make a bar chart showing the number of cells per sample, coloured
by condition (healthy vs cirrhotic). Use `ggplot2`. Hint: start from the `sapply()` output
above and turn it into a data frame with `enframe()`.

```{r cells-barplot}
# Your code here
sapply(seurat_list, ncol) %>%
  tibble::enframe(name = "sample_name", value = "n_cells") %>%
  left_join(sample_meta, by = "sample_name") %>%
  ggplot(aes(x = sample_name, y = n_cells, fill = condition)) +
  geom_col() +
  scale_fill_manual(values = c("healthy" = "#4CAF50", "cirrhotic" = "#F44336")) +
  labs(
    title = "Number of cells per sample",
    x     = "Sample",
    y     = "Cell count",
    fill  = "Condition"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question 3:** Cell counts vary considerably across samples. Is this a problem?
 What biological or technical reasons might explain large differences in cell yield
 between samples?



---

## 4. Merging the A/B Lanes

Three donors had their CD45− fraction split across two sequencing lanes (A and B). These
are the **same biological sample** run across two lanes to increase depth — they must be
merged into one object before QC so they are treated as a single sample throughout.
This is often done in bulk RNA sequencing as well where samples may be multiplexed
across different lanes and you simply merge the reads together. 

▶ **Provided** — this merging logic is roundabout;  run as-is. The actual mechanics
are not the focus, just the reason for why we are merging. 

```{r merge-ab}
# This function finds A/B lane pairs, merges each pair into one Seurat object,
# and leaves single-lane samples unchanged.
merge_ab_lanes <- function(slist, meta) {
  
  # pull() extracts a column as a plain vector (rather than a one-column data frame).
  # We get the names of all lane-A and lane-B samples from the metadata.
  a_names <- meta %>% filter(lane == "A") %>% pull(sample_name)
  b_names <- meta %>% filter(lane == "B") %>% pull(sample_name)
  
  merged_list <- list()
  
  for (i in seq_along(a_names)) {
    a_nm      <- a_names[i]
    b_nm      <- b_names[i]
    
    # Create the merged sample name by removing the "-A" suffix.
    # e.g. "GSM4041151_healthy1_cd45-A" becomes "GSM4041151_healthy1_cd45-"
    merged_nm <- str_replace(a_nm, "-A", "")
    
    # Seurat's merge() combines two or more Seurat objects into one.
    # add.cell.ids prefixes every barcode with "laneA_" or "laneB_".
    merged_list[[merged_nm]] <- merge(
      x            = slist[[a_nm]],
      y            = slist[[b_nm]],
      add.cell.ids = c("laneA", "laneB")
    )
    cat("Merged:", a_nm, "+", b_nm, "->", merged_nm, "\n")
  }
  
  split_names    <- c(a_names, b_names)
  single_samples <- slist[!names(slist) %in% split_names]
  
  # c() on two lists concatenates them into one list
  c(single_samples, merged_list)
}

seurat_list <- merge_ab_lanes(seurat_list, sample_meta)
cat("\nTotal samples after merging:", length(seurat_list), "\n")
```

> **Question 4:** Why do we add `laneA_` and `laneB_` prefixes to the cell barcodes
> when merging? What would happen if we did not?
>
> *Your answer:*

---

## 5. Quality Control

QC in single-cell RNA-seq removes three types of low-quality observations:

1. **Empty droplets** — droplets with no cell; contain only ambient RNA (very low gene count)
2. **Damaged cells** — cells that lost cytoplasmic RNA during dissociation, leaving behind
   mitochondrial RNA (very high % mitochondrial reads)
3. **Doublets** — two cells in one droplet, inflating gene and UMI counts (Section 7)

### 5.1 Computing QC Metrics

We need three metrics per cell: `nFeature_RNA` and `nCount_RNA` are already present.
We need to add `percent.mt`.

Mitochondrial genes in the human genome are named with the prefix `MT-`. The function
`PercentageFeatureSet()` calculates what percentage of a cell's total counts come from
genes matching a given pattern. When a cell is damaged or dying, the outer membrane breaks
down and cytoplasmic mRNA leaks out — but mitochondria (having their own membrane) are
retained longer, so the proportion of mitochondrial reads increases.

Note that the mitochondrial gene naming convention varies by species — in mouse it is
`mt-` (lowercase). Always check for your organism.

Use the PercentageFeatureSet to calculate statistics for mitochondrial gene expression
across all the barcodes / cells. 

```{r qc-metrics}
# We use lapply() here for the same reason as before: it applies the same
# operation to every sample without copy-pasting.
# The anonymous function (function(so) { ... }) is a compact way to write
# a function that is only used in one place.
seurat_list <- lapply(seurat_list, function(so) {
  
  # PercentageFeatureSet() scans every gene name for the pattern "^MT-".
  # The ^ is a regular expression anchor meaning "starts with" --
  # so "^MT-" matches genes whose names begin with MT-, like MT-CO1, MT-ND1.
  # Without the ^, it would also match genes that merely contain "MT-"
  # anywhere in their name, which could cause false positives.
  # The result is stored as a new column "percent.mt" in @meta.data.
  # FILL IN CODE HERE
  so$percent.mt <- PercentageFeatureSet(so, pattern = "^MT-")

  
  
  
  return(so)
})

# Confirm the new column appears in the metadata of the first sample
head(seurat_list[[1]]@meta.data)
```

### 5.2 Visualising QC Metrics

Always visualize distributions **before** filtering so you can make informed decisions.
If you filter first, you cannot see what you removed.

▶ **Provided** — merge samples temporarily for plotting.

```{r pre-qc-merge}
# We create a temporary merged object just so we can plot all samples together.
# merge() here is Seurat's merge function (not base R's merge for data frames).
# add.cell.ids prefixes each cell's barcode with its sample name, so barcodes
# remain unique across samples.
# This object is only for visualisation -- we continue working with the
# per-sample list (seurat_list) for QC filtering.
so_preqc <- merge(
  x            = seurat_list[[1]],   # First seurat object
  y            = seurat_list[-1],    # seurat_list[-1] means "all elements except the first" - so all other seurat objects
  add.cell.ids = names(seurat_list)
)
```

**Your turn** — use `VlnPlot()` to make violin plots of `nFeature_RNA`, `nCount_RNA`,
and `percent.mt`, grouped by sample, in a single column (`ncol = 1`). Set `pt.size = 0`
to hide individual points. Run `?VlnPlot` if you need help with the arguments.

```{r qc-violin, fig.height=10}
# Your code here
VlnPlot(
  so_preqc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  group.by = "sample",
  ncol     = 1,
  pt.size  = 0
) & theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
```

▶ **Provided** — a scatter plot showing the relationship between UMI count and gene count.

```{r qc-scatter}
# FeatureScatter() plots two metadata variables against each other, one cell per point.
# The strong positive correlation between nCount_RNA and nFeature_RNA is expected:
# a cell with more total counts will generally detect more distinct genes.
# Cells that deviate strongly from this trend are suspicious:
#   - Bottom-right (high UMI, low genes): possibly empty droplets or dying cells
#     where a few highly expressed transcripts dominate
#   - Top-right (high UMI, high genes): possibly doublets
FeatureScatter(
  object   = so_preqc,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  group.by = "condition",
  pt.size  = 0.3
) +
  labs(
    title    = "nFeatures vs. nCounts",
    subtitle = "Top-right = possible doublets | Bottom-left = possible empty droplets"
  )
```

**Question 5:** Look at your violin plots. Do the distributions look similar across all
samples? Note any samples that look notably different. What might explain a sample with
a much higher median `percent.mt`?


### 5.3 MAD-Based Adaptive Filtering

A common pattern is applying the same hard cutoffs to every sample — e.g. removing all cells
with `percent.mt > 20%`. The problem is that these thresholds are arbitrary and ignore real
differences between samples. A sample with generally high mitochondrial content (perhaps due
to tissue type or dissociation conditions) would be over-filtered.

A better approach uses the **median absolute deviation (MAD)** to identify outliers relative
to each sample's own distribution. The idea is simple:

```{r mad-explainer}
# Toy example: find the outlier in this set of values
values <- c(500, 520, 480, 510, 495, 505, 490, 2400)

# The median is robust to outliers -- it is not pulled toward the extreme value.
# The MAD measures how spread out the typical values are, also robustly.
cat("Median:", median(values), "\n")
cat("MAD:   ", mad(values), "\n\n")

# A value more than 3 MADs above the median is considered an "outlier".
# This threshold adapts to each sample's own distribution rather than
# using a fixed number that may be too strict or too lenient for a given sample.
upper_limit <- median(values) + 3 * mad(values)
cat("Upper limit (median + 3*MAD):", round(upper_limit), "\n")
cat("Which values are flagged?", values > upper_limit, "\n")
```

The function `isOutlier()` from `scuttle` does this automatically, per sample, and on the
log scale (better for count data that spans orders of magnitude).

Complete the `nmads` argument and the `type` for the mitochondrial filter.
Think carefully: should high mitochondrial percentage be caught by `"lower"` or `"higher"`?

```{r qc-mad}
# We write this as a function (rather than inline code) for two reasons:
# 1. We need to run it on every sample -- wrapping it in a function means
#    we can use lapply() cleanly
# 2. It makes it easy to change nmads in one place and re-run everything
filter_by_mad <- function(so, nmads) {
  
  # Seurat v5 stores data in "layers" which can be split across samples after merging.
  # JoinLayers() consolidates them into a single layer, which is required before
  # converting to a SingleCellExperiment object.
  so  <- JoinLayers(so)
  
  # as.SingleCellExperiment() converts our Seurat object to a different but compatible
  # format used by Bioconductor packages like scuttle. The QC metrics we need
  # (nCount_RNA, nFeature_RNA, percent.mt) are carried over automatically.
  sce <- as.SingleCellExperiment(so)
  
  # isOutlier() computes the MAD-based threshold and returns a logical vector:
  # TRUE for cells that are outliers (to be flagged), FALSE for cells that pass.
  #
  # metric: the vector of values to test (one number per cell)
  # nmads:  how many MADs from the median counts as "too extreme"
  # type:   "lower" flags values that are unusually LOW
  #         "higher" flags values that are unusually HIGH
  # log:    TRUE means the test is done on log(metric), which is more appropriate
  #         for count data that spans several orders of magnitude

  # TODO: USE THIS FUNCTION MULTIPLE TIMES TO TEST VARIOUS QC METRICS FOR ANY OUTLIERS
  
    # nFeature_RNA: flag cells with unusually LOW gene counts (empty droplets)
  #low_features  <- isOutlier(sce$nFeature_RNA,  nmads = nmads, type = "lower",  log = TRUE)
  low_gene  <- isOutlier(sce$nFeature_RNA, nmads = nmads, type = "lower",  log = TRUE)

  
  # nCount_RNA: flag cells with unusually LOW UMI counts (empty droplets)
  #low_counts    <- isOutlier(sce$nCount_RNA,    nmads = nmads, type = "lower",  log = TRUE)
  low_lib   <- isOutlier(sce$nCount_RNA,   nmads = nmads, type = "lower",  log = TRUE)

  
  # percent.mt: flag cells with unusually HIGH mitochondrial % (damaged/dying cells)
  #high_mt       <- isOutlier(sce$percent.mt,    nmads = nmads, type = "higher", log = FALSE)
  high_mt   <- isOutlier(sce$percent.mt,   nmads = nmads, type = "higher", log = FALSE)

  # log = FALSE for percent.mt because it can be 0, and log(0) = -Inf which breaks the calculation
  
  high_gene <- isOutlier(sce$nFeature_RNA, nmads = nmads, type = "higher", log = TRUE)


  
  # The | operator means OR: a cell is discarded if it fails ANY of the checks.
  # A cell only needs to fail one metric to be considered low quality.
  # TODO: CREATE A COLUMN THAT CHECKS IF ANY SINGLE METRIC FAILS
  
  # A cell is discarded if it fails ANY single metric
  discard_flag <- low_lib | low_gene | high_gene | high_mt
  
  # We store each flag as a separate column in @meta.data.
  # This lets us see exactly WHY each cell was flagged, not just that it was.
  # This is useful for troubleshooting and for inspecting borderline samples.
  # TODO: CREATE A NEW COLUMN THAT STORES THE RESULTS FOR EACH OF THE QC METRICS YOU CHOOSE

  # Store each flag separately so we can see WHY a cell was flagged
  so$qc_low_lib    <- low_lib
  so$qc_low_gene   <- low_gene
  so$qc_high_gene  <- high_gene
  so$qc_high_mt    <- high_mt
  so$qc_discard    <- discard_flag
  
  return(so)
}

# lapply() applies filter_by_mad() to every sample in seurat_list.
# The result is a new list of Seurat objects, each with QC columns added.
# We pass nmads as an extra argument -- lapply() forwards it to filter_by_mad().
# Try nmads = 3 first, then revisit in Question 7.
seurat_list <- lapply(
  
  seurat_list,
  filter_by_mad,
  nmads = 3
  
)

```
```{r}
print(seurat_list)
```


▶ **Provided** — summarise how many cells are flagged per sample.

```{r qc-summary}
# Another lapply() pattern: extract a summary from each sample, then
# bind_rows() stacks all the per-sample data frames into one table.

# You'll notice we are simply calling the columns we created above
qc_summary <- lapply(names(seurat_list), function(nm) {
  md <- seurat_list[[nm]]@meta.data
  data.frame(
    sample        = nm,
    total_cells   = nrow(md),
    low_lib       = sum(md$qc_low_lib),
    low_gene      = sum(md$qc_low_gene),
    high_gene     = sum(md$qc_high_gene),
    high_mt       = sum(md$qc_high_mt),
    total_discard = sum(md$qc_discard),
    pct_discard   = round(100 * mean(md$qc_discard), 1)
  )
}) %>% bind_rows()

print(qc_summary)
```

▶ **Provided** — Visualise which cells are flagged.

```{r qc-discard-plot}
# Now that filter_by_mad() has added qc_discard to each sample's metadata,
# we rebuild the merged object so those columns are already present.
# This avoids the barcode alignment problems that arise from trying to transfer
# flags from the per-sample list into a separately-created merged object.
so_preqc <- merge(
  x            = seurat_list[[1]],
  y            = seurat_list[-1],
  add.cell.ids = names(seurat_list)
)

# qc_discard is already in @meta.data -- no extra transfer step needed
# You'll notice the group.by allows us to color by a variable we created in
# the metadata
FeatureScatter(
  object   = so_preqc,
  feature1 = "nCount_RNA",
  feature2 = "percent.mt",
  group.by = "qc_discard",
  pt.size  = 0.3,
  cols     = c("grey80", "firebrick")
) +
  labs(title  = "MAD-flagged cells",
       colour = "Flagged for removal")
```

**Question 6:** Look at the `pct_discard` column. Are some samples losing more cells
than others? At what percentage of cells discarded would you start to worry that something
went wrong with the whole sample?



**Question 7:** We used `nmads = 3`. Go back and re-run the filter with `nmads = 2`.
How many more cells are removed? Now try `nmads = 5`. What is the trade-off you are making
as you move the threshold in each direction? Is there a single correct value?


### 5.4 Applying the Filter

Remove the flagged cells. We want to keep cells where `qc_discard` is `FALSE`.

```{r apply-qc}
# subset() on a Seurat object works similarly to filtering rows in a data frame.
# The subset argument is evaluated against @meta.data -- so qc_discard here
# refers to the column we just added.
# We use lapply() to apply this to every sample, again producing a new list
# where each object contains only the cells that passed QC.

# TODO: FIGURE OUT HOW TO USE THE SEURAT SUBSET TO REMOVE ANY CELLS WHERE qc_discard IS FALSE
seurat_list <- lapply(seurat_list, function(so) {

  
})

cells_after_qc <- sapply(seurat_list, ncol)
cat("Cells remaining after QC:\n")
print(cells_after_qc)
cat("\nTotal:", sum(cells_after_qc), "\n")
```

---


## 6. Doublet Detection

A **doublet** is a droplet containing two cells. It appears as one "cell" with roughly
twice the expected number of genes and UMIs. If not removed, doublets can form spurious
clusters between cell types or inflate gene expression estimates.

**Critical:** doublet detection must be run **per sample, before merging**. If you run
it on the merged object, the algorithm sees cells from many different donors and cell types.
High-expressing cells from one donor can look like doublets of lower-expressing cells from
another, producing many false positives. Running it per sample ensures the algorithm only
compares cells from the same library, where a genuinely high-count cell is more clearly
anomalous.

`scDblFinder()` takes a `SingleCellExperiment` object. The two output
columns you need are `scDblFinder.score` (a continuous score) and `scDblFinder.class`
(the call: `"singlet"` or `"doublet"`). Transfer them to the Seurat metadata.

```{r doublets}
seurat_list <- lapply(seurat_list, function(so) {
  
  # Convert to SingleCellExperiment -- scDblFinder requires this format.
  # JoinLayers() is needed first in Seurat v5 for the same reason as in filter_by_mad().
  so  <- JoinLayers(so)
  sce <- as.SingleCellExperiment(so)
  
  # scDblFinder() simulates artificial doublets by combining random pairs of cells,
  # then trains a classifier to distinguish real cells from these simulated doublets.
  # It adds results directly to the SCE object's cell metadata.
  sce <- scDblFinder(sce = sce)
  
  # Transfer results back to the Seurat metadata.
  # scDblFinder.score: a continuous probability (0-1); higher = more likely a doublet
  # scDblFinder.class: the binary call -- "singlet" or "doublet"
  so$doublet_score <- sce$___   # <-- continuous score column
  so$doublet_class <- sce$___   # <-- singlet/doublet call column
  
  return(so)
})
```

▶ **Provided** — summarise doublet calls and visualise scores. This pattern of
code should be familiar - lapply using an anonymous function which extracts
data from each seurat object and binds it together into a single dataframe to
summarize the entire dataset. 

```{r doublet-summary}
doublet_summary <- lapply(names(seurat_list), function(nm) {
  so <- seurat_list[[nm]]
  data.frame(
    sample       = nm,
    n_cells      = ncol(so),
    n_doublets   = sum(so$doublet_class == "doublet"),
    pct_doublets = round(100 * mean(so$doublet_class == "doublet"), 1)
  )
}) %>% bind_rows()

print(doublet_summary)
```

```{r doublet-plot}
# Merge for plotting only -- same pattern as the pre-QC merge above.
# geom_hline() adds a horizontal reference line at 0.5.
so_predbl <- merge(seurat_list[[1]], y = seurat_list[-1],
                   add.cell.ids = names(seurat_list))

VlnPlot(
  object   = so_predbl,
  features = "doublet_score",
  group.by = "sample",
  pt.size  = 0
) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "firebrick") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(title    = "Doublet scores per sample",
       subtitle = "Cells above the dashed line are called doublets")
```

**Question 9:** The expected doublet rate for 10X Genomics is ~0.8% per 1,000 cells
loaded. Do the rates you observe match this expectation? What are the consequences of
being too aggressive with doublet removal (i.e. a high false positive rate)?



```{r remove-doublets}
seurat_list <- lapply(seurat_list, function(so) {
  subset(
    x      = so,
    subset = doublet_class == "singlet" 
  )
})

cat("Cells after doublet removal:\n")
print(sapply(seurat_list, ncol))
```

---

## 8. Normalisation and Variable Feature Selection

We need to normalise for differences in sequencing depth before comparing cells.
A cell with 10,000 UMIs will appear to express every gene at twice the level of an
otherwise identical cell with 5,000 UMIs — a purely technical difference.

**Log normalisation:** divide each cell's counts by its total UMIs, multiply by 10,000,
then take the natural log. This puts all cells on a comparable scale regardless of how
deeply they were sequenced.

We also select the **highly variable genes (HVGs)** — genes that vary most across cells.
These are the most informative for distinguishing cell types and will be used in PCA.
Genes that are expressed at the same level in every cell carry no information for
clustering and are excluded to reduce noise and computation time.

**Why do this per sample rather than on the merged object?** Normalising per sample
ensures that each sample's own library size distribution is used as the reference.
Identifying variable genes per sample and then taking the union (which Seurat does
automatically at the merge step) avoids genes that are only "variable" because of
technical differences between batches.

Complete the normalisation and variable feature selection. The function
names are `NormalizeData()` and `FindVariableFeatures()`. Run `?NormalizeData` if needed.

```{r normalise}
seurat_list <- lapply(seurat_list, function(so) {
  
  # NormalizeData() normalises the raw counts and stores the result in the
  # "data" layer of the Seurat object (the "counts" layer is kept unchanged).
  # normalization.method = "LogNormalize": divide by total counts, multiply by
  #   scale.factor, then take log1p (log(x + 1), the +1 avoids log(0)).
  # scale.factor = 10000: normalises to "counts per 10,000", often called CP10K.
  #   This makes values interpretable across cells with different sequencing depths.
  so <- NormalizeData(
    object               = so,
    normalization.method = "___",   # <-- 
    scale.factor         = ___      # <-- 
  )
  
  # FindVariableFeatures() identifies genes that vary more across cells than
  # would be expected by chance given their mean expression level.
  # selection.method = "vst": variance-stabilising transformation -- fits a curve
  #   to the mean-variance relationship and selects genes above the curve.
  # nfeatures: how many top variable genes to select.
  so <- FindVariableFeatures(
    object           = so,
    selection.method = "vst",
    nfeatures        = ___   # <-- 
  )
  
  return(so)
})

# VariableFeatures() retrieves the list of selected HVGs from a Seurat object.
top10 <- head(VariableFeatures(seurat_list[[1]]), 10)
cat("Top 10 highly variable genes:\n")
print(top10)
```

**Question 10:** Look at the top variable genes. Do any look like known immune cell
markers? Does that make sense for the CD45+ fraction? Are there any genes you would
not expect to see here?



---

## 9. Merging All Samples

Now that each sample has been QC-filtered, doublet-removed, and normalised, we combine
them into a single Seurat object for joint analysis.

▶ **Provided** — merge all samples and verify the metadata.

```{r merge-all}
# merge() with multiple objects: the first goes in x, the rest go in y as a list.
# add.cell.ids prefixes every barcode with the sample name to keep them unique.
# merge.data = TRUE carries over the normalised data (the "data" layer) from
# each sample, not just the raw counts. Without this, we would need to re-normalise
# after merging.
#
# Note: after merging in Seurat v5, counts and data are stored as separate named
# layers (e.g. "counts.sample1", "data.sample1"). We will join these with
# JoinLayers() before ScaleData() below.
so_merged <- merge(
  x            = seurat_list[[1]],
  y            = seurat_list[-1],
  add.cell.ids = names(seurat_list),
  merge.data   = TRUE
)

so_merged

cat("\nTotal cells:", ncol(so_merged), "\n")
cat("Total genes:", nrow(so_merged), "\n")
```


---

## 10. Pre-processing Before Integration

We scale the data and run PCA to reduce dimensionality before integration.

**Scaling** transforms each gene to have mean 0 and variance 1 across all cells. This
is necessary before PCA because PCA is sensitive to the scale of each variable —
a highly expressed gene with large absolute values would dominate the PCA axes simply
due to scale, not because it is biologically informative.

**PCA** (principal component analysis) finds the linear combinations of genes that
explain the most variation across cells. Instead of working with ~18,000 genes
(most of which are noisy), we reduce the data to the top 50 principal components,
which capture the major axes of biological variation much more efficiently.

▶ **Provided** — join layers, scale, and run PCA.

```{r scale-pca}
# JoinLayers() consolidates the per-sample layers created by merge() into
# single "counts" and "data" layers. This is required in Seurat v5 before
# ScaleData() and most downstream steps.
so_merged <- JoinLayers(so_merged)

# ScaleData() centres and scales each gene across all cells.
# features = rownames(so_merged) scales all genes, not just the HVGs.
# Scaling all genes is slower but necessary if you want to visualise
# non-HVG genes in heatmaps or feature plots later.
so_merged <- ScaleData(
  object   = so_merged,
  features = rownames(so_merged)
)

# RunPCA() performs PCA on the scaled HVG expression matrix.
# npcs = 50: compute the top 50 principal components.
#   We will decide how many to actually use based on the elbow plot.
# verbose = FALSE: suppress the per-PC output (too much text)
so_merged <- RunPCA(
  object  = so_merged,
  npcs    = 50,
  verbose = FALSE
)
```

Produce an elbow plot using `ElbowPlot()`. This shows the standard
deviation explained by each PC. Look for where the curve flattens — beyond that point,
additional PCs capture mostly noise rather than real biological signal.

```{r elbow}
# Your code here

```

Set `n_pcs` based on your elbow plot.

```{r set-npcs}
# Based on the elbow plot, I will use this many PCs:
n_pcs <- ___   # <-- your choice


```

### 10.1 Inspect Batch Structure Before Integration

Before correcting for batch effects, we visualise the data in 2D using UMAP to understand
what the dominant sources of variation are. If cells separate primarily by donor rather
than by cell type or condition, integration is clearly needed.

**UMAP** (Uniform Manifold Approximation and Projection) takes the high-dimensional PCA
coordinates and projects them into two dimensions for visualisation. Cells that are similar
in gene expression end up close together; dissimilar cells end up far apart.

▶ **Provided** — build a cell neighbourhood graph and run UMAP on the unintegrated PCA.

```{r umap-unintegrated}
# FindNeighbors() builds a k-nearest neighbour graph in PCA space.
# Each cell is connected to its most similar neighbours.
# This graph is used by both FindClusters() (in Part 2) and RunUMAP() here.
# dims = 1:n_pcs: use the first n_pcs principal components as the input space.
so_merged <- FindNeighbors(
  object    = so_merged,
  dims      = 1:n_pcs,
  reduction = "pca"
)

# RunUMAP() takes the neighbour graph and produces a 2D embedding.
# reduction.name = "umap.unintegrated" saves this UMAP under a custom name
# so we can compare it to the Harmony UMAP later without overwriting it.
so_merged <- RunUMAP(
  object         = so_merged,
  dims           = 1:n_pcs,
  reduction      = "pca",
  reduction.name = "umap.unintegrated"
)
```

Make three UMAP plots of the unintegrated data, colouring cells by
`"donor"`, `"condition"`, and `"fraction"` respectively. Use `DimPlot()` with
`reduction = "umap.unintegrated"`. Arrange them in a column using `/` from `patchwork`.

```{r umap-unint-plots, fig.height=14}
# Your code here

```

**Question 11:** Which variable — donor, condition, or fraction — most strongly
separates cells in the unintegrated UMAP? Is this what you expected?

Now think about what to do with this: which sources of variation should we **remove**
(because they are technical batch effects) and which should we **preserve** (because
they reflect real biology we want to study)?




## 11. Decisions Log

Fill in this table before moving to Part 2. There are no uniquely correct answers, but
you should be able to justify every choice.

| Step | Value used | Cells removed | Your reasoning |
|------|-----------|---------------|----------------|
| `min.cells` | | — | |
| `min.features` | | — | |
| `percent.mt` pattern | `^MT-` | — | Mitochondrial genes in humans |
| MAD `nmads` threshold | | | |
| Doublet removal | — | | |
| HVGs selected | | — | |
| Number of PCs | | — | |
| **Total cells remaining** | | | |

---

## 12. Save for Part 2

▶ **Provided** — save the cleaned, integrated object.

```{r save}
# saveRDS() saves a single R object to a file. readRDS() in Part 2 will
# restore it exactly as it is now, including all reductions and metadata.
saveRDS(object = so_merged, file = "liver_cirrhosis_part1.rds")

cat("Saved: liver_cirrhosis_part1.rds\n")
cat("Cells:", ncol(so_merged), "\n")
cat("Genes:", nrow(so_merged), "\n")
cat("Reductions:", paste(names(so_merged@reductions), collapse = ", "), "\n")
```

In Part 2 we will cluster these cells, identify cell types using known marker genes,
and run pseudobulk differential expression analysis between healthy and cirrhotic liver.

---

## Session Info

```{r session-info}
sessionInfo()
```
