---
title: "Human Liver Cirrhosis scRNA-seq — Part 2: Clustering & Marker Gene Analysis"
subtitle: "Ramachandran et al. 2019 | GSE136103"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

## Learning Objectives

By the end of this exercise you will be able to:

1. Run the full Seurat clustering pipeline — variable features, scaling, PCA, UMAP, graph, clusters — within each fraction
2. Use `clustree` to choose a clustering resolution and aid in justifying the choice
3. Run `FindAllMarkers()` and interpret the key output columns
4. Localise the TREM2+CD9+ macrophage subpopulation described in the paper
5. Annotate clusters in both fractions using canonical marker genes

**Estimated time:** 1-2 hours

---

## Background

Ramachandran et al. sorted each liver biopsy into two fractions:

- **CD45+** — haematopoietic cells: macrophages, Kupffer cells, T cells, NK cells,
  B cells, monocytes
- **CD45−** — non-haematopoietic cells: hepatocytes, cholangiocytes, endothelial
  cells, hepatic stellate cells

The authors clustered each fraction **separately** using standard Seurat PCA-based
clustering, with resolutions tuned to the complexity of each compartment. We replicate
that approach here.

For each fraction we run a fresh pipeline from scratch: variable
feature selection → scaling → PCA → UMAP → neighbour graph → clustering. The PCA
embeddings are used consistently throughout.


## 0. Libraries

▶ **Provided**

```{r libraries}
library(Seurat)
library(tidyverse)
library(patchwork)
library(clustree)

set.seed(42)
```

---


```{r split-fractions}

# the data can be found at /projectnb/bf530/materials/cd45+_liver_cirrhosis.rds

so_pos <- readRDS('/projectnb/bf530/materials/cd45+_liver_cirrhosis.rds')

so_pos

# An object of class Seurat 
# 23043 features across 32626 samples within 1 assay 
#Active assay: RNA (23043 features, 2000 variable features)
# 21 layers present: counts.1, counts.2, counts.4, counts.5, counts.7, counts.8, counts.9, counts.11, counts.13, counts.14, data.1, data.2, data.4, data.5, data.7, #data.8, data.9, data.11, data.13, data.14, scale.data
# 2 dimensional reductions calculated: pca, umap.unintegrated
```


```{r}
# In Seurat v5, subsetting can leave data split across named layers.
# JoinLayers() merges them back into single "counts" and "data" layers,
# which is required before FindVariableFeatures() and ScaleData().
so_pos <- JoinLayers(so_pos)

so_pos

# An object of class Seurat 
# 23043 features across 32626 samples within 1 assay 
# Active assay: RNA (23043 features, 2000 variable features)
# 3 layers present: data, counts, scale.data
# 2 dimensional reductions calculated: pca, umap.unintegrated
```


---

## 2. Full Pre-processing Pipeline Per Fraction

We run the complete pipeline independently within each fraction. Working per-fraction
ensures that the highly variable genes, the PCA axes, and the UMAP layout all reflect
variation *within* the cd45+ cells. 

*Provided** — a single function encapsulates the whole pipeline so the same steps
are applied consistently.

```{r pipeline-fn}
# run_pipeline() applies the full Seurat pre-processing and clustering-preparation
# steps to one fraction object:
#
#   FindVariableFeatures  — identify the most informative genes for this fraction
#   ScaleData             — center and scale for PCA
#   RunPCA                — reduce to top n_pcs principal components
#   RunUMAP               — 2-D layout for visualisation (stored as umap_name)
#   FindNeighbors         — build KNN graph on PCA embeddings
#
# FindClusters() is called separately in Section 3 so the resolution sweep
# can be done without re-running the expensive steps above.

run_pipeline <- function(so, n_hvg = 2000, n_pcs = 30,
                         umap_name = "umap") {
  so <- FindVariableFeatures(so, nfeatures = n_hvg,      verbose = FALSE)
  so <- ScaleData(so,          features = rownames(so),  verbose = FALSE)
  so <- RunPCA(so,             npcs     = n_pcs,         verbose = FALSE)
  so <- RunUMAP(so,
                dims           = 1:n_pcs,
                reduction      = "pca",
                reduction.name = umap_name,
                verbose        = FALSE)
  so <- FindNeighbors(so,
                      dims      = 1:n_pcs,
                      reduction = "pca",
                      verbose   = FALSE)
  return(so)
}
```

Inspect the elbow plots first to choose `n_pcs` for each fraction,
then run the full pipeline.

```{r elbow-plots, fig.height=4}
# Run PCA with 50 PCs temporarily to see the full elbow curve.
# We only need ScaleData and RunPCA for this — the rest of the pipeline
# will be run properly below once you have chosen n_pcs.

so_pos_tmp <- so_pos %>%
  FindVariableFeatures(nfeatures = 2000, verbose = FALSE) %>%
  ScaleData(features = rownames(so_pos), verbose = FALSE) %>%
  RunPCA(npcs = 50, verbose = FALSE)



ElbowPlot(so_pos_tmp, ndims = 50) + ggtitle("CD45+ elbow")
```

```{r run-pipeline}
# Set n_pcs based on the elbow plots above.
# n_pcs controls variable feature selection, scaling, PCA, UMAP, AND the
# neighbour graph — it is the single parameter that flows through the whole pipeline.
n_pcs_pos <-  20 # <-- your choice for CD45+


so_pos <- run_pipeline(so_pos, n_pcs = n_pcs_pos, umap_name = "umap_pos")

```


---

## 3. Resolution Sweep and clustree

`FindClusters()` takes the KNN graph built by `FindNeighbors()` and finds communities
within it using the Louvain algorithm. The `resolution` parameter controls granularity:
higher values produce more clusters. We test a range, store all results in `@meta.data`,
and use `clustree` to choose.

### 3.1 CD45+ Resolution Sweep

Complete the `resolution` argument in the loop.

```{r res-sweep-pos}
resolutions_pos <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.8, 1.0)

for (res in resolutions_pos) {
  so_pos <- FindClusters(
    object     = so_pos,
    resolution = res,
    verbose = FALSE
                        # Add in the parameters in your loop for resolution
  )

cat ("CD45+ | res", res, "->",
  length(unique(so_pos@meta.data[[paste0("RNA_snn_res.", res)]])),
  "clusters\n")
}
```

```{r clustree-pos, fig.height=9, fig.width=8}
# Each column = one resolution. Node size = cell count.
# Edges show where cells move as resolution increases.
# Look for: stable nodes that persist without gaining cells from many parents
# (robust populations), and the resolution at which a small new cluster first
# separates cleanly from a larger macrophage node (likely the TREM2+ population).

# so_pos@meta.data will show you the cluster assignments for each resolution
# Each separate resolution is a new column added with the cluster assignments

clustree(so_pos, prefix = "RNA_snn_res.") # This is the prefix of the column
```

```{r}
so_pos@meta.data
```


Generate two clustree plots now highlighting the expression of the TREM2 and CD9
genes - the markers of the scar associated macrophages identified by the paper.

```{r}
clustree(so_pos, prefix = "RNA_snn_res.", node_colour = "TREM2",
         node_colour_aggr = "mean", exprs = "data")

clustree(so_pos, prefix = "RNA_snn_res.", node_colour = "CD9",   
         node_colour_aggr = "mean", exprs = "data")
```


```{r choose-res-pos}
chosen_res_pos  <-  0.1             # Choose a resolution based on the clustree results
cluster_col_pos <- paste0("RNA_snn_res.", chosen_res_pos)
Idents(so_pos)  <- cluster_col_pos

```

Note:

When no reference annotation is available, Clustree interpretation becomes centered
around stability: run FindClusters across a range of resolutions 
(typically 0.1 to 1.5 in steps of 0.1) and treat the tree as a map of where the 
data wants to split. A resolution is a reasonable choice if its nodes each have 
a single dominant parent — meaning cells are flowing cleanly from one cluster 
into two daughter clusters as resolution increases, rather than being pulled from
multiple sources simultaneously. Multiple incoming arrows to a node signal 
instability: the algorithm is forcing a split that the underlying data does not
strongly support. The practical rule is to choose the lowest resolution at which
biologically interpretable structure first appears, defined here not by marker 
gene identity but by internal coherence — stable nodes, clean splits, and cluster
sizes large enough to be meaningful given your total cell count. 


If you have known reference cell types that you expect to find, you can also
overlay their expression information onto the clustree plot to help you visualize
at what resolution does the marker become stable in a cluster or various clusters. 

---

## 4. UMAPs at Chosen Resolutions

Create three UMAP plots using `reduction = "umap_pos"` and use `group.by` to overlay
the `condition`, your chosen resolution (RNA_snn_res.X.X), and the `sample` information.


```{r umap-pos, fig.height=5}
# CD45+ 
# Your code here
DimPlot(so_pos, reduction = "umap_pos", group.by = "condition")
DimPlot(so_pos, reduction = "umap_pos", group.by = "RNA_snn_res.0.1")
DimPlot(so_pos, reduction = "umap_pos", group.by = "sample")
```

Do cluster boundaries align with condition (healthy vs cirrhotic)?
What would it mean biologically if a cluster were almost exclusively from one
condition? Compare what you see here with the unintegrated UMAP from Part 1 —
are donor effects visible within either fraction?


---

## 5. TREM2 and CD9 in the CD45+ Fraction

The TREM2+CD9+ population is in the CD45+ fraction. Visualise these genes directly
before running the formal marker analysis.

### 5.1 Feature Plots

Create a `FeaturePlot()` for `TREM2` and `CD9`.
`reduction = "umap_pos"`, `ncol = 2`, `order = TRUE`.

```{r featureplot-trem2-cd9, fig.height=5}
# Your code here

```

### 5.2 Violin Plots Across CD45+ Clusters

Create a `VlnPlot()` for `TREM2` and `CD9`. `pt.size = 0`, `ncol = 1`.

```{r vlnplot-trem2-cd9, fig.height=8}
# Your code here

```

---

## 6. Marker Gene Analysis

### 6.1 CD45+ Markers

*Provided** — Allow several minutes. Most marker gene analyses use a simple
wilcoxon rank sum test comparing the expression in every individual cluster vs.
the expression of all other clusters. The p-values from these tests are often
inflated but for the purposes of marker gene analysis, they are often reliable
enough.

```{r markers-pos}
markers_pos <- FindAllMarkers(
  object          = so_pos,
  only.pos        = TRUE,
  min.pct         = 0.25,
  logfc.threshold = 0.25,
  verbose         = TRUE
)

write.csv(markers_pos, "markers_cd45pos.csv", row.names = FALSE)
cat("CD45+ marker rows:", nrow(markers_pos), "\n")
head(markers_pos)
```


Key columns:

| Column | Meaning |
|--------|---------|
| `avg_log2FC` | Log2 fold-change of the cluster vs all other cells |
| `pct.1` | Fraction of cells *in* the cluster expressing the gene |
| `pct.2` | Fraction of cells *outside* the cluster expressing the gene |
| `p_val_adj` | BH-adjusted p-value |
| `cluster` | Which cluster this row describes |


### 6.2 Top Markers Per Cluster

Extract out the top 5 markers by `avg_log2FC` per cluster, filtered to
`p_val_adj < 0.05`.

```{r top-markers}
top5_pos <- markers_pos %>%
  filter() %>%
  group_by() %>%
  slice_max(order_by = , n = 5) %>%
  ungroup()

```

### 6.3 Marker Heatmaps

Use `DoHeatmap()` for each fraction using the genes in `top5_pos` /
`top5_neg`.

```{r heatmap-pos, fig.height=10, fig.width=12}
# CD45+ heatmap
# Hint: pass top5_pos$gene to the features argument
# Add + theme(axis.text.y = element_text(size = 7)) if labels overlap
# Your code here

```

---

## 7. Deep Dive: The TREM2+CD9+ Macrophage Cluster

### 7.1 Markers of the Candidate Cluster

Fill in the cluster number from earlier that represents the
putative TREM2 - CD9 Macrophage cluster

```{r trem2-cluster-markers}
candidate_cluster <- "___"   # <-- from earlier

trem2_cluster_markers <- markers_pos %>%
  filter(cluster == candidate_cluster, p_val_adj < 0.05) %>%
  arrange(desc(avg_log2FC))

head(trem2_cluster_markers, 20)
```

Beyond `TREM2` and `CD9`, what are the top markers for this cluster?
Ramachandran et al. report `GPNMB`, `LGMN`, and `SPP1` as additional markers,
are they present in your list? Look up any genes you do not recognize.


### 7.2 Dot Plot of Canonical TREM2+ Markers

Use `DotPlot()` on `so_pos` for the panel below. Add `+ coord_flip()`.

```{r dotplot-trem2, fig.height=7}
trem2_panel <- c(
  "TREM2", "CD9",
  "GPNMB", "LGMN", "SPP1",
  "C1QA", "C1QB",
  "MARCO", "CD163",
  "FCGR3A",
  "CD14", "S100A8"
)

# Your code here

```

### 7.3 Condition Enrichment

Calculate and plot the proportion of each CD45+ cluster made up of
cirrhotic vs healthy cells as a stacked bar chart.

```{r condition-proportions, fig.height=5}


```

Is the candidate cluster enriched in cirrhotic samples? Is this
consistent with the paper's finding? State one caveat about interpreting cell
proportions from samples that may differ in cell yield or sequencing depth.


---

## 8. Cell Type Annotation

### 8.1 CD45+ Annotation

| Cell type | Canonical markers |
|-----------|-------------------|
| Kupffer cells | `C1QA`, `C1QB`, `MARCO`, `CD163` |
| TREM2+ macrophages | `TREM2`, `CD9`, `GPNMB`, `SPP1` |
| Classical monocytes | `CD14`, `FCGR3A`, `S100A8`, `S100A9` |
| CD8 T cells | `CD3D`, `CD8A`, `GZMB` |
| CD4 T cells | `CD3D`, `CD4`, `IL7R` |
| NK cells | `GNLY`, `NKG7`, `KLRD1` |
| B cells | `CD19`, `MS4A1`, `CD79A` |
| Plasma cells | `IGHG1`, `MZB1`, `SDC1` |

Edit `annotation_map_pos` to match your cluster count.

```{r annotate-pos}
annotation_map_pos <- c(
  "0" = "Unknown",
  "1" = "Unknown",
  "2" = "Unknown",
  "3" = "Unknown",
  "4" = "Unknown",
  "5" = "Unknown",
  "6" = "Unknown",
  "7" = "Unknown"
  # add or remove rows to match your cluster count
)

so_pos$cell_type <- annotation_map_pos[
  as.character(so_pos@meta.data[[cluster_col_pos]])
]

cat("CD45+ cell type counts:\n")
print(table(so_pos$cell_type))
```

### 8.2 Annotated UMAPs

Create labelled UMAP for the cd45+ cells with `group.by = "cell_type"` and
`label = TRUE`. Arrange side by side with `patchwork`.

```{r annotated-umaps, fig.height=5}
# Your code here

```

---

## 9. Save the object

```{r save}
saveRDS(so_pos,
        paste0("liver_cirrhosis_cd45pos_res", chosen_res_pos, ".rds"))

```

---

## 10. Optional

If you finish all of this, feel free to also examine the CD45- fraction using
the same principles as above.

### CD45− Annotation

| Cell type | Canonical markers |
|-----------|-------------------|
| Hepatocytes | `ALB`, `APOB`, `CYP3A4`, `TTR` |
| Cholangiocytes | `KRT19`, `KRT7`, `EPCAM`, `CFTR` |
| ACKR1+ endothelial | `ACKR1`, `PECAM1`, `VWF` |
| PLVAP+ endothelial | `PLVAP`, `PECAM1`, `CD34` |
| Hepatic stellate cells | `ACTA2`, `COL1A1`, `LRAT`, `PDGFRB` |
| Portal fibroblasts | `PDGFRA`, `COL1A1`, `DES` |


## Session Info

```{r session-info}
sessionInfo()
```
